---
title: "5: Data Exploration"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
fig_width: 5
fig_height: 2.5
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Import and explore datasets in R
2. Graphically explore datasets in R
3. Apply data exploration skills to a real-world example dataset

## Opening discussion: why do we explore our data?

Why is data exploration our first step in analyzing a dataset? What information do we gain? How does data exploration aid in our decision-making for data analysis steps further down the pipeline?

## Import data and view summaries

```{r, message = FALSE}
# 1. Set up your working directory
getwd()

# 2. Load packages
library(tidyverse)

# 3. Import datasets
USGS.flow.data <- read.csv("./Data/Processed/USGS_Site02085000_Flow_Processed.csv") #when data stored as .csv, dates are saved as CHARACTERS rather than dates --> must tell R that dates are dates!

View(USGS.flow.data)
# Alternate option: click on data frame in Environment tab

colnames(USGS.flow.data)
str(USGS.flow.data)
dim(USGS.flow.data)

# Check our date column
class(USGS.flow.data$datetime)

USGS.flow.data$datetime <- as.Date(USGS.flow.data$datetime, format = "%Y-%m-%d") #changing class of datetime column into dates 
class(USGS.flow.data$datetime) #now class is a date objet!

```


## Visualization for Data Exploration

```{r}
#remember when you begin data exploration, to keep in mind the class of the values in your dataset --> is it categorical? numerical? this is important to know before beginning visualization!
```


Although the `summary()` function is helpful in getting an idea of the spread of values in a numeric dataset, it can be useful to create visual representations of the data to help form hypotheses and direct downstream data analysis. Below is a summary of the useful types of graphs for data exploration.

Note: each of these approaches utilize the package "ggplot2". We will be covering the syntax of ggplot in a later lesson, but for now you should familiarize yourself with the functionality of what each command is doing.

### Bar Chart (function: geom_bar) > good for categorical data

Visualize count data for categorical variables. 

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data, aes(x = discharge.mean.approval)) +
  geom_bar() ##ggplot() is function, first factor is data that you will plot, aes() is to set the data aeshetics/specific portion of data to plot and asking for the mean, geom_bar() is saying what type of plot

ggplot(na.omit(USGS.flow.data), aes(x = discharge.mean.approval)) +
  geom_bar() #this is tweeking the data so that no NAs are in the graphed data
```

### Histogram (function: geom_histogram) > for value intervals

Visualize distributions of values for continuous numerical variables. What is happening in each line of code? Insert a comment above each line.

```{r, fig.height = 3, fig.width = 4}
#
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean))

#binwidth argument specific to histogram, tells R to go beyond set binwidth to x units from the mean
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), binwidth = 10)

#this sets the number of bins you want in your chart; generally use between 5-20 bins
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), bins = 20)

#scale_x_continuous zooms in on where the data is within histogram (sets the min and max on the x-axis)
ggplot(USGS.flow.data, aes(x = discharge.mean)) +
  geom_histogram(binwidth = 10) + 
  scale_x_continuous(limits = c(0, 500))
  
#
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = gage.height.mean)) +
  scale_x_continuous(limits = c(0,12))

```
### Frequency line graph (function: geom_freqpoly)

An alternate to a histogram is a frequency polygon graph (distributions of values for continuous numerical variables). Instead of displaying bars,  counts of continuous variables are displayed as lines. This is advantageous if you want to display multiple variables or categories of variables at once.

```{r, fig.height = 3, fig.width = 4}
#can change color of line, lty is the line type
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean), bins = 50) +
  geom_freqpoly(aes(x = gage.height.min), bins = 50, color = "red") +
  geom_freqpoly(aes(x = gage.height.max), bins = 50,  lty = 2) +
  scale_x_continuous(limits = c(0, 10))

#theme() can add a legend and adjust position; within color, it looks for the unique values and color each record by the unique values

#can set color to a categorical value and set two different categories on the same plot
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean, color = gage.height.mean.approval), bins = 50) +
  scale_x_continuous(limits = c(0, 10)) +
  theme(legend.position = "top")

```
### Box-and-whisker plots (function: geom_boxplot, geom_violin) > summary of our data

A box-and-whisker plot is yet another alternative to histograms (distributions of values for continuous numerical variables). These plots consist of: 

* A box from the 25th to the 75th percentile of the data, called the interquartile range (IQR).

* A bold line inside the box representing the median value of the data. Whether the median is in the center or off to one side of the IQR will give you an idea about the skewness of your data.

* A line outside of the box representing values falling within 1.5 times the IQR. 

* Points representing outliers, values that fall outside 1.5 times the IQR. 

An alternate option is a violin plot, which displays density distributions, somewhat like a hybrid of the box-and-whiskers and the frequency polygon plot.

```{r, fig.height = 3, fig.width = 4}
#middle line --> median of the records -show distribution of values
#box is showing the quantiles
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean.approval, y = gage.height.mean)) 

#takes different slices of data and looking at distribution; cut-width takes distribution of data and shows distribution of values within the requested data
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean, y = discharge.mean, group = cut_width(gage.height.mean, 1)))

#adding a little more distribution (see range of values and distribution within range); width is the amount of data at each slice
ggplot(USGS.flow.data) +
  geom_violin(aes(x = gage.height.mean.approval, y = gage.height.mean), 
              draw_quantiles = c(0.25, 0.5, 0.75))
```

### Scatterplot (function: geom_point)
Visualize relationships between continuous numerical variables.

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data) +
  geom_point(aes(x = discharge.mean, y = gage.height.mean))

ggplot(USGS.flow.data) +
  geom_point(aes(x = datetime, y = discharge.mean))
```

Question: under what circumstances would it be beneficial to use each of these graph types (bar plot, histogram, frequency polygon, box-and whisker, violin, scatterplot)?

> Answer: 

## Ending discussion

What did you learn about the USGS discharge dataset today? What separate insights did the different graph types offer?
> Answer: 


How can multiple options for data exploration inform our understanding of our data?

> Answer: 


Do you see any patterns in the USGS data for the Eno River? What might be responsible for those patterns and/or relationships?

> Answer: 


