---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Meilin Chan"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(agricolae)

ntl.lter.chemphys <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

ntl.lter.chemphys$sampledate <- as.Date(ntl.lter.chemphys$sampledate, 
                                        format = "%m/%d/%y")

#2

my_theme <- mytheme <- theme_bw(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(my_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature recorded during July does not change with depth across all lakes.
Ha: The mean lake temperature recorded during July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
library(lubridate)

ntl.lter.july <- ntl.lter.chemphys %>%
  mutate(Month = month(sampledate)) %>%
  filter(Month == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  filter(!is.na(temperature_C)) 

print(ntl.lter.july)

#5
scatter.ntl.lter.july <- ggplot(ntl.lter.july, 
                                aes(depth, temperature_C)) +
  geom_point() +
  geom_smooth(method = lm, aes(color = "red"))+
  theme(legend.position = "none") +
  xlab("Depth (meters)") +
  ylab("Temperature in C")+
  ylim(0,35)

print(scatter.ntl.lter.july)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Based on the figure, there are deviations between data points and the linear regression. Although generally it looks that temperature does decrease with depth, it does not do so in a linear fashion. The temperature looks to decrease exponentially as depth increases, meaning the temperature of the water decreases more rapidly with depth than would by seen in a linear relationship. Overall, it looks like temperature and depth do not have a strong linear relationship.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
depth.temp.regression <- lm(data = ntl.lter.july, 
                            temperature_C ~ depth)
summary(depth.temp.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: From the linear regression, overall we can reject the null hypothesis since p-value < 0.05, which means that there is a relationship between temperature and depth. The regression also has a R-squared value of 0.7387 showing that while there is variability in temperature as there are changes in depth (ie: the temperatures do not have a very strong linear relationship with depth), there is a relationship and it does follow a trend. Depth explains 74% of the variability of temperature. There is a residual standard error of 3.835 on 9726 degrees of freedom - this means that there are 9726 variable/free points that do not align with the linear regression, and the larger residual standard error shows there is more variation in the relationship than we would want to see in a linear relationship. So although the p-value shows that temperature and depth are related, the relationship between the two is variable and does not follow a strong linear trend To calculate temperature predicted for every 1m change in depth, you multiply the depth by -1.95 and then add 21.96 - this will give you the estimated temperature in celsius. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
aic.ntl.lter <- lm(data = ntl.lter.july, 
                   temperature_C ~ year4 + daynum + depth)

step(aic.ntl.lter) #lowest AIC number is when all 3 variables are within the model

#10
summary(aic.ntl.lter)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests are all 3 that we had input - year4, daynum, and depth. Removing any of these variables would lead to a worse model than the original one. This model has an R-sqaured of 0.741 compared to the single linear model using temperature's R-squared of 0.738. There was a slight improvement, but not by very much. The relationship is still not very linear. Our multiple regression model does not explain a lot of the observed variance as including additional variables within the model did not reveal a much stronger linear relationship.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
aov.lake.temp <- aov(data = ntl.lter.july, temperature_C ~ lakename)
summary(aov.lake.temp)

lm.lake.temp <- lm(data = ntl.lter.july, temperature_C ~ lakename)
summary(lm.lake.temp)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer:  Using aov(), we find that the p-value < 0.05 for the lakes, showing that there is a significiant difference in mean temperatures among the lakes. Observing the lm(), we can see the amount of variation of the means of each lake compared to the mean of Central Long Lake in the Estimate column, visually showing us that each lake mean is different. We can also see that the p-value is the same (p-value < 0.05) - aka: we reject the null hypothesis that all the means are the same. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
library(tidyverse)
library(ggridges)
library(viridis)
library(RColorBrewer)
library(colormap)
scat.lm.lake.temp <- ggplot(ntl.lter.july, 
                            aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5, aes(color = lakename)) +
  geom_smooth(method = lm, se = FALSE, color = "black") +
  facet_wrap(vars(lakename), nrow = 3) +
  xlab("Depth (meters)") +
  ylab("Temperature in Celsius") +
  ylim(0,35) +
  theme(legend.position = "none")
  
print(scat.lm.lake.temp)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(aov.lake.temp)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: 
>statistically significant mean (p-value < 0.05) - Central Long Lake, Crampton Lake, East Long Lake, Hummingbird Lake, Tuesday Lake, West Long
>NOT statistically significant mean (p-value > 0.05) - Paul Lake, Ward Lake

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could run a t-test!


