---
title: "Assignment 5: Data Visualization"
author: "Meilin Chan"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(cowplot)

PP_Nutrients <- read.csv("../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = FALSE)

NN_Litter <- read.csv("../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
class(PP_Nutrients$sampledate)
PP_Nutrients$sampledate <- as.Date(PP_Nutrients$sampledate, 
                                   format = "%Y-%m-%d")
class(PP_Nutrients$sampledate)

class(NN_Litter$collectDate)
NN_Litter$collectDate <- as.Date(NN_Litter$collectDate, 
                                 format = "%Y-%m-%d")
class(NN_Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
A05_theme <- theme_bw(base_size = 12) +
  theme(axis.text = element_text(color = "black"), legend.position = "bottom")
theme_set(A05_theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
library(ggplot2)

NTL.LTER.phos.Peter <- ggplot(subset(PP_Nutrients, lakename == "Peter Lake"), 
                              aes(x = po4, y = tp_ug)) +
  geom_point(color = "purple", size = 0.8, shape = "square")+
  xlim(0, 45) +
  ylim(0, 150) +
  geom_smooth(method = lm, color = "black") +
  xlab("Phosphate") +
  ylab("Phosphorus") +
  ggtitle("Peter Lake")
print(NTL.LTER.phos.Peter)

NTL.LTER.phos.Paul <- ggplot(subset(PP_Nutrients, lakename == "Paul Lake"), 
                             aes(x = po4, y = tp_ug)) +
  geom_point(color = "red", size = 0.8) +
  xlim(0, 35) +
  ylim(0, 150) +
  geom_smooth(method = lm, color = "black") +
  xlab("Phosphate") +
  ylab("Phosphorus") +
  ggtitle("Paul Lake")
print(NTL.LTER.phos.Paul)


NTL.LTER.phos <- plot_grid(NTL.LTER.phos.Paul, NTL.LTER.phos.Peter, 
                           nrow = 2, align = 'hv', rel_heights = c(3,3))
print(NTL.LTER.phos)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5
PP_Nutrients$month <- as.factor(PP_Nutrients$month)

NTL.temp <- ggplot(PP_Nutrients, aes(x = month, y = temperature_C)) +
  geom_boxplot(aes(color = lakename)) +
  xlab("Month") +
  ylab("Temperature (Celsius)") +
  labs(color = "Lake Name")

print(NTL.temp)

NTL.TP <- ggplot(PP_Nutrients, aes(x = month, y = tp_ug)) +
  geom_boxplot(aes(color = lakename)) +
  xlab("Month") +
  ylab("ug") +
  xlim("5","6","7","8","9")

print(NTL.TP)


NTL.TN <- ggplot(PP_Nutrients, aes(x = month, y = tn_ug)) +
  geom_boxplot(aes(color = lakename)) +
  xlab("Month") +
  ylab("ug") +
  xlim("5","6","7","8","9")

print(NTL.TN)

NTL.legend <- get_legend(NTL.temp)

print(plot_grid(NTL.legend)) #only shows legend - no axes

NTL.three <- plot_grid(NTL.temp + theme(legend.position = "none"),
                       NTL.TP + theme(legend.position = "none"),
                       NTL.TN + theme(legend.position = "none"),
                       nrow = 1, 
                       align = 'hv',
                       labels = c("Temp", "TP", "TN"))
print(NTL.three) 


NTL.three.leg <- plot_grid(NTL.three, NTL.legend,nrow = 2, 
                           rel_heights = c(3,0.3))

print(NTL.three.leg)


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The temperature of both Peter Lake and Paul Lake increase as the 
season shifts from winter to summer and then decreases again as the season 
shifts back into autumn. TP in Peter Lake increases in ug as the season shifts 
into summer and then autumn. TP in Paul Lake decreases as the season shifts 
into summer but then increases again as it becomes autumn. TN in Peter Lake 
increases and peaks as winter shifts into summer (around August), but then 
decreases once the season starts to roll into autumn. TN in Paul Lake seems 
to remain at a steady ug level until around September when it increases. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
class(NN_Litter$collectDate)
NN_Litter$collectDate <- as.Date(NN_Litter$collectDate)


needles.dry.point <- ggplot(subset(NN_Litter, functionalGroup == "Needles")) +
  geom_point(aes(x = collectDate, y = dryMass, 
                 color = nlcdClass), alpha = 0.7) +
  theme(legend.position = "bottom") +
  xlab("Collection Year") +
  ylab("Needles Dry Mass") +
  labs(color = "NLCD Class") +
  scale_color_manual(labels = c("Evergreen Forest", "Grassland Herbaceous", 
                                "Shrub Scrub"),
                     values = c("dark green", "blue", "orange"))
                      
print(needles.dry.point)

#7

pointfacet.needles.dry <- ggplot(subset(NN_Litter, functionalGroup == "Needles")) +
  geom_point(aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  theme(legend.position = "none") +
  xlab("Collection Year") +
  ylab("Needles Dry Mass") +
  facet_grid(~ nlcdClass)

print(pointfacet.needles.dry)
  

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why?

> Answer: Plot 7 is more effective. By separating NLCD classes into 3 facets,
we are able to better see the overall spread of amount of needle dry mass 
collected over the timespan of the data. From Plot 7, we can see that across 
all NLCD classes, that most of the dry mass collected over the course of the 
time period was below 10 units of dry mass. When we observe Plot 6, the plot is difficult to read as some of the points overlap with each other and it is 
difficult to see the full distribution of each NLCD Class during each collection 
periods. 
