---
title: "Assignment 09: Data Scraping"
author: "Meilin Chan"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()

library(tidyverse)
library(rvest)
library(dataRetrieval)
library(lubridate)
library(viridis)
library(tinytex)

my_theme <- theme_classic() +
  theme(legend.position = "bottom",
        legend.text = element_text(color = "black"))
theme_set(my_theme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
website.20 <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * MAX Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- website.20 %>%
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>%
  html_text()
water.system.name #Durham
  
pswid <- website.20 %>%
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>%
  html_text()
pswid #03-32-010
  
ownership <- website.20 %>%
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
  html_text()
ownership #Municipality

max.withdrawals.mgd <- website.20 %>%
  html_nodes('th~ td+ td') %>%
  html_text()
max.withdrawals.mgd #dim 1:12; 1st value = 36.0100

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4
df_maxwith <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", 
                                     "June","Oct", "March", "July", "Nov", 
                                     "April", "Aug", "Dec"),
                         "Year" = rep(2020, 12),
                         "Avg_Max_Withdrawals" = as.numeric(max.withdrawals.mgd),
                         "PSWID" = pswid,
                         "Water_System" = rep(water.system.name),
                         "Ownership" = rep(ownership)
                         )

df_maxwith <- df_maxwith %>%
  mutate("Date" = my(paste(Month, "-", Year))) %>%
  select(Date, Avg_Max_Withdrawals, 
         PSWID, Water_System, Ownership)

df_maxwith <- df_maxwith[order(df_maxwith$Date, decreasing = FALSE),]


#5
maxdaily_plot <- ggplot(df_maxwith, aes(x=Date, y=Avg_Max_Withdrawals)) +
  geom_line()+
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2020 Water Withdrawals in",
                     water.system.name,ownership),
       y="Avg. Max. Withdrawals (mgd)",
       x="Date")
  
plot(maxdaily_plot)



```

6. Note that the PSWID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PSWID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.

scrape.NCDEQ <- function(the_year,pswid_num){
  the_url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",
                    pswid_num,"&year=",the_year)
  #link website
  the_website <- read_html(the_url)
  
  #locating elements
  system.name.fun <- the_website %>%
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>%
  html_text()
  
  pswid.fun <- the_website %>%
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>%
  html_text()
  
  ownership.fun <- the_website %>%
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
  html_text()
  
  max.withdrawals.fun <- the_website %>%
  html_nodes('th~ td+ td') %>%
  html_text()
  
  month <- c("Jan", "May", "Sep", "Feb", "June",
             "Oct", "March", "July", "Nov", "April", "Aug", "Dec")
  
  the_df <- data.frame("Month" = month,
                         "Year" = rep(the_year, 12),
                         "Avg_Max_Withdrawals" = as.numeric(max.withdrawals.fun),
                         "PSWID" = pswid.fun,
                         "Water_System" = rep(system.name.fun),
                         "Ownership" = rep(ownership.fun)
                         )
  
  the_df <- the_df %>%
  mutate("Date" = my(paste(Month, "-", Year))) %>%
  select(Date, Avg_Max_Withdrawals, PSWID, Water_System, Ownership)
  
  the_df <- the_df[order(the_df$Date, decreasing = FALSE),]
  
  return(the_df)
  
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
NDEQ_15 <- scrape.NCDEQ('2015','03-32-010')

summary(NDEQ_15)

ggplot(NDEQ_15, aes(x=Date, y =Avg_Max_Withdrawals))+
  geom_line()+
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Water Withdrawals in",
                     water.system.name,ownership),
       y="Avg. Max. Withdrawals (mgd)",
       x="Date")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
ashe.2015 <- scrape.NCDEQ('2015','01-11-010')

ggplot() + 
  geom_line(data=ashe.2015, 
            aes(x = Date, y =Avg_Max_Withdrawals, color = "Asheville")) +
  geom_line(data=NDEQ_15, 
            aes(x = Date, y =Avg_Max_Withdrawals, color = "Durham")) +
  labs(title = "2015 Water Withdrawals in NC Municipalities",
       y="Avg. Max. Withdrawals (mgd)",
       x="Date") +
  theme(legend.position = "bottom")
  

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9

ashe.2015 <- scrape.NCDEQ('2015','01-11-010')
ashe.2016 <- scrape.NCDEQ('2016','01-11-010')
ashe.2017 <- scrape.NCDEQ('2017','01-11-010')
ashe.2018 <- scrape.NCDEQ('2018','01-11-010')
ashe.2019 <- scrape.NCDEQ('2019','01-11-010')

ashe.15.16 <- full_join(ashe.2015, ashe.2016)
ashe.15.17 <- full_join(ashe.15.16, ashe.2017)
ashe.15.18 <- full_join(ashe.15.17, ashe.2018)
ashe.15.19 <- full_join(ashe.15.18, ashe.2019)

ggplot(ashe.15.19, aes(x=Date, y =Avg_Max_Withdrawals))+
  geom_line()+
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Asheville Water Withdrawals from 2015-2019",
       y="Avg. Max. Withdrawals (mgd)",
       x="Date")

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

> Yes, Asheville's water usage over time has generally increased from 2015 - 2019